---
title: "R workshop_4"
author: "Jackie Maud"
date: "21 January 2019"
output: html_document
---
# Creating objects in R
```{r arithmetic}

3+5



#assign value to object

weight_kg<-55

# to convert to pounds

weight_lb <-2.2 * weight_kg


sqrt(weight_kg)

# rounds off to required number sigfig)

round(3.14159) #default is 3 sigfig
round(3.14159, digits=2)  
#or
round(3.14159, 2)
```
###Vectors and data types

Some basic data types in R

```{r}
weight_g<-c(50, 60, 65, 82)

animals<-c("mouse", "rat", "dog")



```
# Different types of vector (i.e. atomic - only one type of data)

*numeric
*character
*logical (TRUE or FALSE)
*factors (categorical)
*dates

A vector is a data structure  (has a linear structure)

Other data structures:
  
*lists
*data frames
*matrices

## Data frames

```{r}
library(tidyverse)

#utils::  tells us which source to get package from

#Dowmload data onto computer, need next function to read the data

download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv")

library(here) #This package makes working directories and file paths made easy

#

surveys<-read_csv(here("read_data", "portal_data_joined.csv"))

surveys

#columns of data frame are vectors

#data frame as vectors pf equal length
#matrix can have columns of different length

str(surveys)
dim(surveys)
nrow(surveys)
summary(surveys)

```
###Indexing and subsetting data frames

First using square brackets []

Square brackets are great for defining 

Do restart and reload (green down arrow)


```{r set up, warning = FALSE}

#First define the row coordinate, and then the column, Also write row and then column

surveys[1,1]

surveys[1,6]

#Defining only which element we want will return a data frame

surveys[1]

surveys[1:3, 7]

# Give us all rows and columns except column 7

surveys[, -c(1:5)]

```
## Data manipulation  (tidyverse)

Key functions for data manipulation

*select(): subsetting columns
*filter(): subsets of rows based on conditions
*mutate(): create new columns, based on info from other columns
*group_by(): creates groups based on categorical data
*summarize(): create summary stats on grouped data
*arrange(): sort results
*count(): gives a count of discrete values
```{r}
```{r}
select(surveys, plot_id, species_id, weight)

select(surveys, -record_id)

#use == for logical statements

filter(surveys, year==1995)

filter(surveys, year==1995,
       species_id=="NL")
```


##Pipes

Pipe: ctrl-shift-m

###Write multiple arguments in a sentence

```{r}
surveys %>% 
  filter(weight<5) %>% 
  select(species_id, sex, weight)

surveys_sml<-surveys %>%
filter(weight<5) %>% 
  select(species_id, sex, weight)

```


#Challenge 1

All data for 1995 and year, sex and weight


```{r}
surveys_jlm<- surveys %>%
  filter(year==1995) %>% 
  select(year, sex, weight)

surveys_jlm
```


```{r}
surveys %>%
  mutate(weight_kg = weight/1000,  #creates new column
         weight_kg2 = weight_kg*2)  # creates new column based on new column
         

surveys %>% 
  drop_na(weight)  # drops NAs from specified column
  
  
  
```

## Challenge 2

Only species_id. New column called hindfoot_half: contains half foot length values. Also, no NAs in hindfoot_half column. Values <30 


```{r}
surveys_jlm<-surveys %>%
  drop_na(hindfoot_length) %>% 
  mutate(hindfoot_half=hindfoot_length/2) %>% 
  filter(hindfoot_half <30) %>% 
  select(species_id, hindfoot_half)
```


```{r}
surveys %>%     #find mean weight by sex and ignore NAs in weight
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE))
  
surveys %>%     #find mean weight by species and sex and ignore NAs in weight column
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE)) 
  
surveys %>%     #find mean and min weight by species and sex and ignore NAs in weight column
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE))
            
            
 surveys %>%     #find mean and min weight by species and sex,ignore NAs in weight column and sort by min weight in ascending order
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>% 
   arrange(min_weight)        # default is ascending, for descending need to add function arrange(desc(min_weight))         
  

  
```

```{r}
surveys %>% 
  count(sex)
  

```
## Challenge 3

1. How many animals were caught in each plot type surveyed?

2. Use group_by and summarize to find mean, min & max of hindfoot_length for each species (using species_id). Also, add no. observations (hint: see ?n)

3. What was the heaviest animal measured in each year? Return year, genus, species_id and weight

```{r}
#1.

surveys %>% 
  count(plot_type)
  
  
  
#2.

surveys %>% 
  group_by(species_id) %>%
    summarize(mean_length = mean(hindfoot_length, na.rm = TRUE),
            min_length = min(hindfoot_length, na.rm = TRUE),
            max_length = max(hindfoot_length, na.rm = TRUE), 
            n = n())
  
#3.


max_weights <-surveys %>% 
  drop_na(weight) %>% 
  group_by(year) %>% 
  filter(weight==max(weight)) %>% 
  select(year, genus, species, weight) %>% 
  arrange(year)
  
max_weights  

```
## Export our data


```{r}
write_csv(max_weights, here("write_data", "max_weights.csv"))
```


